---
output: html_document
---
Do California Students at Charter Schools Score Better on the SAT and ACT?
========================================================
## by Eric Gordon
```{r global_options, include=FALSE}
knitr::opts_chunk$set (fig.width=10, fig.height=6, fig.align='center',
                      fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r  packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.
library(maps)
library(zipcode)
library(ggplot2)
library(mapdata)
library(reshape2)
library(dplyr)
library(plyr)
library(tidyr)
library(gridExtra)
library(ggthemes)
library(GGally)
```

```{r   Load_the_Data}
# Load the Data
setwd("/Users/ericgordo/Documents/Udacity/P4_EDA_Course_Materials/P4/P4_EDA_CA_Schools/Data")
#import Files as Dataframes
act<-read.csv("act15.csv", stringsAsFactors = FALSE)
sat<-read.csv("sat15.csv", stringsAsFactors = FALSE)
schools<-read.csv("pubschls.csv", stringsAsFactors = FALSE)
dropouts<-read.csv("filesdropouts.txt", sep="\t", stringsAsFactors = FALSE)

# Work for the Data I want! 
#act data Cleaning
act <- act %>% 
  subset (rtype=="S") %>%  
  subset (select=-c (rtype,Enroll12,
                   year,dname,cname,sname)) %>% #Columns to Drop
  subset(AvgScrEng!='*' & !is.na(AvgScrEng)) %>%   #Drop Columns with NO Values
  #Change Data Types For Numeric Columns 
  transform (AvgScrEng=as.numeric(AvgScrEng),
            AvgScrMath=as.numeric(AvgScrMath),
            AvgScrRead=as.numeric(AvgScrRead), 
            AvgScrSci=as.numeric(AvgScrSci),
            NumGE21=as.numeric(NumGE21),
            PctGE21=as.numeric(PctGE21)) 

#SAT data cleaning
sat<- sat %>%
  subset (sat$rtype=="S") %>% #Only schools, not District Totals
  subset (select=-c(rtype,enroll12,dname,cname,sname)) %>%  #Columns to Drop
  subset (AvgScrMath!='*' & !is.na(AvgScrMath)) %>%   #Drop Columns with NO Values
   #Change Data Types For Numeric Columns
  transform (AvgScrRead=as.numeric (AvgScrRead),
            AvgScrMath=as.numeric (AvgScrMath),
            AvgScrWrite=as.numeric (AvgScrWrite),
            PctGE1500=as.numeric (PctGE1500),
            NumGE1500=as.numeric (NumGE1500))
sat$AvgSum <- sat$AvgScrRead + sat$AvgScrWrite + sat$AvgScrMath #Add An Average Combined Score 

#schools data cleaning
schools <- schools%>%
  subset (schools$StatusType=="Active") %>% #Only Active Schools
  subset (select=c (CDSCode,County,District,
                    School,City,
                    Charter,Magnet,
                    Latitude,Longitude, 
                    EdOpsName)) # Columns to keep

# Dropout Data
dropouts <- dropouts %>%
  subset (select=c(CDS_CODE,DTOT,ETOT))
dropouts <- aggregate(data=dropouts,
                      cbind(ETOT, DTOT)~CDS_CODE, sum)

#Final DataFrame Creation
school.df <- merge (act, sat, by = "cds", all=TRUE)
school.df <- school.df %>% 
  merge (dropouts, by.x="cds", by.y="CDS_CODE")%>%
  merge (schools, by.x = "cds", by.y="CDSCode")
setwd("/Users/ericgordo/Documents/Udacity/P4_EDA_Course_Materials/P4/P4_EDA_CA_Schools")
```

**This Analysis is jsut a portion of the entire investigation into this data. To read or access files from this full investigation, click [here](https://github.com/ericgordo/P4_EDA_CA_Schools)**

This analysis compares schools across the state of California. Specifically, it looks at the percent of students who score above the national average on the SAT and ACT standardized tests. The average score of the SAT is a 1500, and the average score on the ACT test is 21.

### The Data Investigated
The data in this analysis is the collection of publicly accessible files from the California Department of Education (CDE). This data combines the average test scores from 1,252 California public schools from the 2014-2015 school year.

## Percent of Students who Score Above Average on Each Test

Each point on the chart below represents a school. The X axis position of each point represents the percent of students at the school who scored above average on the ACT, while the Y axis position represents the percent of students scoring above average on the SAT. Thus the schools in the top right corner have more students who score above average on both tests, while the bottom left corner represents schools in which the smallest percent of students perform above average on each test. The orange points represent traditional public schools, and the blue dots represent charter schools.

```{r  Plot_One}
ggplot(data =school.df, 
                   aes(x=PctGE21,y=PctGE1500, color= Charter))+
  geom_point ( alpha=2/3, size=3/4) + 
  geom_hline (yintercept =50, linetype=3) +
  geom_vline (xintercept = 50, linetype=3)+
  scale_color_manual ( values=c("sienna2", "#0000CC"))+
  #geom_smooth (method="lm",se=FALSE, size=1, color="red")+
  labs(color="Charter") +
  xlab ("% Students Scoring Above ACT 21")+
  ylab ("% Students Scoring Above SAT 1500") 

```

The graph above shows the vast difference between schools in the percent of students performing above average on both tests. However there seems to be a disparity between the charter schools and traditional public schools. Let's investigate some stats.

## Further Statistics

```{r charter_stats}
#Getting Stats to Seperate
both <- subset (school.df,
                !is.na(PctGE21) & !is.na(PctGE1500)) 
q1 <- subset (both, 
              PctGE21<50 & PctGE1500<50)
print("Charter Schools on Graph Above")
table(both$Charter)

print("SAT Percent Breakdown")
tapply(school.df$PctGE1500, school.df$Charter, summary)

print("ACT Percent Breakdown")
tapply(school.df$PctGE21, school.df$Charter, summary)

charters <- subset(both, both$Charter=='Y')
non_charters <- subset(both, both$Charter=='N')
```

The outputs above show us several things. First, that there are 125 charter schools on the graph above, and 945 traditional public schools. Also from the “SAT Percent Breakdown”, we see that at the average public school 40% of students score above average on the SAT, while only 33% of students at the average charter schools score above average. Similarly for the ACT, 53% of students perform above average at every public school, compared to just 39.6% at charter schools. We will investigate this further to see if we can draw any statistically significant conclusions from this data.

## Statistical Tests for Significance

We will run a statistical test to see if the mean percent of students who score above average on the SAT and ACT is statistically different between public schools and private schools. That is we will test the following question.

*Is there a significant difference in the percentage of students who score above average on these tests in public vs. charter schools?*

$H_o =$ *There is no difference in the percent of students who score above average on this test*

$H_a =$ *There is a difference in the percent of students who score above average on this test*

```{r  echo=FALSE, Charter_Stats_Test_data_Split}
#ACT
x1 <- charters$PctGE21
y1 <- non_charters$PctGE21

#SAT
x2 <- charters$PctGE1500
y2 <- non_charters$PctGE1500
```

```{r  echo=TRUE, Charter_Stats_Test}
#"ACT Statistical Difference"
wilcox.test(y1, x1, conf.int=TRUE)

#"SAT Statistical Difference"
wilcox.test(x2, y2, conf.int=TRUE)
```

The results above show that there is statistically significant difference in the percent of students who score above average in both tests at public schools compared to charter schools. This can be seen in the small p-values from both tests above. The Wilcoxon-Mann-Whitney test was used because none of the distributions were normal.

## Conclusion

It is interesting that for the 2014-2015 school year, there was a statistically significant difference between the number of students who score above average on both the SAT and ACT tests. This may be surprising, or shocking to many, however there may be further questions worth investigating to see why this result may happen. It is possible that other variables in these schools like the size of schools has a large effect on these results. For the  scope of this investigation, I will leave this analysis here, and recommend that further investigations look into reasons why this result may have occurred. 

# Work Cited
All data used in this analysis found at:
http://www.cde.ca.gov/ds/

```{r   Plot_Three}
usa <- map_data ("state")
ca <- subset (usa, usa$region=="california")

#Basic Map for California
CA.MAP <- ggplot()+ 
  geom_polygon(data=ca, aes(x=long, y=lat),fill="Grey")+ 
  coord_fixed(1.3)+
  theme(panel.background =element_rect(fill ="white"))

CA.ACT <- CA.MAP + 
  geom_point(data=school.df,
             aes(x=Longitude,y=Latitude, colour=PctGE21),size=3/10)+
  scale_color_gradientn(
    colours = colorRampPalette(c("Red", "Blue"))(10))

CA.ACT + labs (title="% of Students Above Avg. (ACT)")
```
